---
title: "Cloud access, versioning, and quality assurance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloud access, versioning, and quality assurance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hfutils)
```

This vignette covers the three responsibilities that make `hfutils` the trusted
base layer of the stack: **authenticated cloud access**, **dataset versioning &
provenance**, and **staged invariant checks**.

## Authenticated cloud access

`lynker_spatial_auth()` provisions (or refreshes) an OAuth2 token and wires it
into the libraries you name. With GDAL configured, any `as_ogr()` /
`sf::read_sf()` call can read directly from cloud storage via a GDAL virtual
file system; with DuckDB configured, the bearer token is attached as a secret.

```{r, eval = FALSE}
# Interactive browser flow; sets GDAL_HTTP_BEARER for the session
token <- lynker_spatial_auth(libs = "gdal")

as_ogr("/vsis3/lynker-spatial/hydrofabric/v2.2/conus/conus_nextgen.gpkg",
       "divides")

# DuckDB over HTTP, with the token attached to the connection
con <- duckdb_connection(extensions = "httpfs")
tbl_http("https://lynker-spatial.s3.amazonaws.com/v20.1/flowpaths.parquet",
         conn = con)
```

## Dataset versioning & provenance

`gpkg_set_version()` stamps a dataset version into the standard GeoPackage
metadata tables — distinct from the GeoPackage *specification* version in
`PRAGMA user_version`, which is left untouched. It records a machine-readable
packed integer (`Mmmpp`), a human-readable semver string, an optional SPDX
license, and an optional JSON provenance record. It is idempotent: re-stamping
replaces prior entries rather than accumulating duplicates.

```{r}
library(sf)

f <- tempfile(fileext = ".gpkg")
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)[1:2, ]
st_write(nc, f, "divides", quiet = TRUE)

gpkg_set_version(
  f, "2.2.0",
  provenance = list(software = "hydrofabric", vpuid = "09"),
  license    = "ODbL-1.0"
)

str(gpkg_get_version(f))
```

`int_version` (`20200` here) is the value GDAL surfaces as
`GPKG_METADATA_ITEM_1`, convenient for machine comparison across releases.

```{r, include = FALSE}
unlink(f)
```

## Staged invariant checks

`hf_check_invariants()` is the shared QA gate used at each stage of the build
pipeline. Calling it *between* steps means a silent failure (a duplicate id, a
dangling `toid`, an orphan divide, a raster ring artifact) halts the run at the
step that created it — not 200 lines later during visual inspection.

It supports four stages — `"refactored"`, `"reconciled"`, `"aggregated"`, and
`"ngen"` — each expecting stage-specific inputs. Here is the topology half of a
`refactored`-stage check on a clean toy network:

```{r}
refactored <- data.frame(
  flowpath_id   = c("101", "102", "103"),
  flowpath_toid = c("102", "103", "0")
)

res <- hf_check_invariants("refactored", refactored = refactored)
res$ok
```

By default any failed check throws (`strict = TRUE`); set `strict = FALSE` to
downgrade failures to warnings while still surfacing them. A dangling downstream
pointer is caught:

```{r, error = TRUE}
broken <- data.frame(
  flowpath_id   = c("101", "102"),
  flowpath_toid = c("102", "999")   # 999 does not exist
)

hf_check_invariants("refactored", refactored = broken)
```

The `aggregated` and `ngen` stages additionally accept `coverage = TRUE` to run
the (more expensive) check that each flowpath lies within its assigned
catchment.
