---
title: "Network properties: accumulation and hydrosequence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network properties: accumulation and hydrosequence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hfutils)
```

A hydrofabric flow network is a **directed acyclic graph (DAG)**: every feature
(`id`) points to exactly one downstream feature (`toid`), and terminal
features point to `0` (or `NA`). `hfutils` provides two topology primitives that
operate on this structure — both are character-safe, so identifiers such as
`"fp-123"` or scientific-notation strings round-trip cleanly.

## Downstream accumulation

`accumulate_downstream()` propagates a per-feature attribute (drainage area,
incremental length, anything additive) downstream, summing all upstream
contributions at confluences. It runs a single topological sort followed by one
O(E) edge pass, so it scales to continental networks.

Consider two headwaters (`1`, `2`) joining at `3`, which flows to the outlet
`4`:

```{r}
df <- data.frame(
  flowpath_id   = c("1", "2", "3", "4"),
  flowpath_toid = c("3", "3", "4", "0"),
  area          = c(1.0, 2.0, 0.5, 0.0)
)

accumulate_downstream(df, attr = "area")
```

Feature `3` accumulates `1.0 + 2.0 + 0.5 = 3.5`, and the outlet inherits the
full basin total. The column names default to `flowpath_id` / `flowpath_toid`
but can be overridden via the `id` and `toid` arguments to match any network
table (including reference-fabric `id`/`toid`).

A cyclic network is rejected rather than silently producing wrong totals:

```{r, error = TRUE}
cyclic <- data.frame(
  flowpath_id   = c("1", "2"),
  flowpath_toid = c("2", "1"),
  area          = c(1, 1)
)

accumulate_downstream(cyclic, attr = "area")
```

## Hydrosequence

`get_hydroseq()` assigns a total topological ordering to the network. Larger
values are upstream; the value decreases monotonically as you move toward the
outlet, which makes it a convenient sort key for downstream traversal.

```{r}
df$hydroseq <- get_hydroseq(df)
df[order(-df$hydroseq), c("flowpath_id", "flowpath_toid", "hydroseq")]
```

## In practice

Both functions accept a data frame, tibble, or `sf` object. A typical pattern
reads a layer lazily, filters to a VPU, materializes it, and accumulates:

```{r, eval = FALSE}
library(dplyr)

da <- as_ogr("conus_nextgen.gpkg", "flowpaths") |>
  filter(vpuid == "01") |>
  st_as_sf() |>
  accumulate_downstream(attr = "areasqkm")
```

See `vignette("reading-and-writing")` for the I/O side of that pipeline.
