---
title: "Network properties: accumulation, ordering, and coding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network properties: accumulation, ordering, and coding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hfutils)
```

A hydrofabric flow network is a **directed acyclic graph (DAG)**: every feature
(`id`) points to exactly one downstream feature (`toid`), and terminal
features point to `0` (or `NA`). `hfutils` provides a family of topology
functions that operate on this structure, all are character-safe, so
identifiers such as `"fp-123"` or scientific-notation strings round-trip
cleanly, and each returns a vector aligned to the input rows.

## Downstream accumulation

`accumulate_downstream()` propagates a per-feature attribute (drainage area,
incremental length, anything additive) downstream, summing all upstream
contributions at confluences. It runs a single topological sort followed by one
O(E) edge pass, so it scales to continental networks.

Consider two headwaters (`1`, `2`) joining at `3`, which flows to the outlet
`4`:

```{r}
df <- data.frame(
  flowpath_id   = c("1", "2", "3", "4"),
  flowpath_toid = c("3", "3", "4", "0"),
  area          = c(1.0, 2.0, 0.5, 0.0)
)

accumulate_downstream(df, attr = "area")
```

Feature `3` accumulates `1.0 + 2.0 + 0.5 = 3.5`, and the outlet inherits the
full basin total. The column names default to `flowpath_id` / `flowpath_toid`
but can be overridden via the `id` and `toid` arguments to match any network
table (including reference-fabric `id`/`toid`).

A cyclic network is rejected rather than silently producing wrong totals:

```{r, error = TRUE}
cyclic <- data.frame(
  flowpath_id   = c("1", "2"),
  flowpath_toid = c("2", "1"),
  area          = c(1, 1)
)

accumulate_downstream(cyclic, attr = "area")
```

## Hydrosequence

`get_hydroseq()` assigns a total topological ordering to the network. Larger
values are upstream; the value decreases monotonically as you move toward the
outlet, which makes it a convenient sort key for downstream traversal.

```{r}
df$hydroseq <- get_hydroseq(df)
df[order(-df$hydroseq), c("flowpath_id", "flowpath_toid", "hydroseq")]
```

## Derived network attributes

The rest of the family builds the standard NHDPlus-style routing attributes on
top of the same topological pass. Take a small basin, a mainstem `4 -> 3 -> 1`
with a tributary `2 -> 1` draining to the outlet `1`:

```{r}
net <- data.frame(
  flowpath_id   = c("1", "2", "3", "4"),
  flowpath_toid = c("0", "1", "1", "3"),
  lengthkm      = c(4, 2, 3, 5)
)

net$arbolate  <- accumulate_downstream(net, attr = "lengthkm")   # arbolate sum
net$order     <- get_streamorder(net)                            # Strahler order
net$levelpath <- get_levelpath(net, weight = "arbolate")         # mainstem grouping
net$pathlen   <- get_pathlength(net, length = "lengthkm")        # distance to outlet
net$level     <- get_streamlevel(net, levelpath = "levelpath")   # stream level
net
```

- **`get_streamorder()`**, Strahler order: leaves are `1`, and order increases
  only where two equal-order streams meet (so the confluence at `1` is order 2).
- **`get_levelpath()`**, groups reaches into continuous mainstems, following the
  largest-`weight` contributor at each confluence (here the arbolate sum, so
  `1`, `3`, `4` share a level path and the tributary `2` is its own).
- **`get_pathlength()`**, distance along the network from each reach's outlet to
  the network terminus (`0` at the outlet, summing downstream lengths going up).
- **`get_streamlevel()`**, how many level-path steps a reach is from the
  terminus: the mainstem is `1`, the tributary `2`.

### Pfafstetter codes

`get_pfafstetter()` assigns hierarchical [Pfafstetter](https://en.wikipedia.org/wiki/Pfafstetter_Coding_System)
basin codes. It needs total drainage area, a hydrosequence, and level paths
precomputed, then subdivides each basin's mainstem by its four largest
tributaries, recursing `max_level` digits deep:

```{r, eval = FALSE}
# on a real fabric this is accumulated catchment area; `arbolate` stands in here
net$total_da_sqkm <- net$arbolate
net$topo_sort     <- get_hydroseq(net)
net$pfaf          <- get_pfafstetter(net, max_level = 2)
```

The result is most meaningful on a full basin, where each digit narrows the
location within the drainage hierarchy. On a four-reach toy network there are
no tributaries large enough to subdivide, so every code comes back `NA`.

## Upstream queries and partitioning

`upstream_index()` assigns each feature a *nested set* pair, `upstream_id` (its
depth-first pre-order position) and `num_upstreams` (how many features sit
strictly above it). Together these turn "everything upstream of X" into an
integer range test rather than a graph traversal:

```{r}
idx <- upstream_index(net)
cbind(net["flowpath_id"], idx)
```

Everything upstream of a feature with `upstream_id == u` and
`num_upstreams == k` is exactly the half-open range `(u, u + k]`:

```{r}
u <- idx$upstream_id[net$flowpath_id == "1"]
k <- idx$num_upstreams[net$flowpath_id == "1"]
net$flowpath_id[idx$upstream_id > u & idx$upstream_id <= u + k]
```

Because the walk expands the largest-upstream branch first, a mainstem stays
contiguous. Note that `upstream_id` is build-specific: it changes whenever the
topology changes, so it is an index, not a persistent key.

For a hydrofabric whose flowpaths route through nexuses, `hf_upstream_index()`
resolves the `flowpath -> nexus -> flowpath` hops first and then indexes the
resulting flowpath graph. `write_hydrofabric()` calls it for you, so a written
GeoPackage carries the index already.

`merge_groups()` builds on the same pre-order to cut the network into
*contiguous* runs, breaking wherever a feature does not flow directly into its
predecessor or (with `order`) wherever the given order column changes. Each
group is a contiguous `upstream_id` range, so a size-budgeted partitioner can
merge whole groups and always end up with complete sub-networks:

```{r}
merge_groups(net, order = "levelpath")
```

The mainstem `4 -> 3 -> 1` forms one group; the tributary `2` is its own.

## In practice

Every function above accepts a data frame, tibble, or `sf` object. A typical
pattern reads a layer lazily, filters to a VPU, materializes it, and
accumulates:

```{r, eval = FALSE}
library(dplyr)
library(sf)

da <- as_ogr("conus_nextgen.gpkg", "flowpaths") |>
  filter(vpuid == "01") |>
  st_as_sf() |>
  accumulate_downstream(attr = "areasqkm")
```

See `vignette("reading-and-writing")` for the I/O side of that pipeline.
