---
title: "Reading and writing a hydrofabric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and writing a hydrofabric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hfutils)
```

`hfutils` is the I/O base layer of the hydrofabric stack. It offers three
complementary access patterns:

1. **`read_hydrofabric()` / `write_hydrofabric()`** — eager, convenience round
   trips for the common `divides` + `flowpaths` realization.
2. **`as_ogr()`** — a lazy, dplyr-compatible view over *any* GDAL/OGR vector
   source (local or cloud), materialized only on `collect()` / `st_as_sf()`.
3. **GeoParquet** (`st_read_parquet()` / `st_write_parquet()`) — columnar,
   cloud-native storage with GeoPandas-compatible metadata.

## A round trip

`read_hydrofabric()` auto-detects flowpath and divide layers by name, and can
harmonize or reproject coordinate reference systems on read. Here we stage a
tiny two-layer GeoPackage and read it back:

```{r}
library(sf)

# Stand-in geometries (the North Carolina sample shipped with sf)
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)[1:5, ]

tmp <- tempfile(fileext = ".gpkg")
write_hydrofabric(
  list(divides = nc, flowpaths = st_cast(nc, "MULTILINESTRING")),
  outfile = tmp,
  verbose = FALSE
)

hf <- read_hydrofabric(tmp, verbose = FALSE)
names(hf)
nrow(hf$divides)
```

`read_hydrofabric()` also accepts in-memory `sf` objects directly (skip the
GeoPackage), a `realization` argument to read only `"divides"` or
`"flowpaths"`, and a `crs` argument to transform outputs on the way out.

```{r}
fp_only <- read_hydrofabric(tmp, realization = "flowpaths", crs = 5070,
  verbose = FALSE)
st_crs(fp_only$flowpaths)$epsg
```

## Lazy access with `as_ogr()`

`as_ogr()` returns a lazy table you can manipulate with dplyr verbs; nothing is
read until you `collect()` (data frame) or `st_as_sf()` (spatial). This is the
efficient way to pull a single VPU out of a continental file without loading the
whole layer:

```{r, eval = FALSE}
library(dplyr)

as_ogr("conus_nextgen.gpkg", "divides") |>
  filter(vpuid == "01") |>
  select(divide_id, areasqkm) |>
  st_as_sf()
```

Because GDAL backs the connection, the same call works against cloud objects via
GDAL virtual file systems — see `vignette("cloud-and-versioning")` for
authenticated `/vsis3/` access.

## GeoParquet

For analytical workflows, write to GeoParquet (single file) or a partitioned
dataset:

```{r, eval = FALSE}
st_write_parquet(hf$divides, "divides.parquet")
back <- st_read_parquet("divides.parquet")
```

```{r, include = FALSE}
unlink(tmp)
```
